---
title: "Qs 5 Rmd 190398965"
author: "Victor Min"
date: "3/24/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q5. Use the available variables to construct a model that predicts delays

## Which Variables would be useful in predicting delays?
## CRSDepTime(scheduled departure time), CRSArrTime(scheduled arrival time), CRSElapsedtime(scheduled time taken), Distance
## Since the dataset is very big, I will only be using one year's worth of data, 2007
```{r}
## Computation times will be much longer here as I went for accuracy in results
memory.limit(45000)
gc()
library(glmnet)
library(caTools)
library(corrplot)

seven <- read.csv("2007.csv", header = TRUE)
Delay <- seven[,-c(15,16)]
Delay[sapply(Delay, is.character)] <- lapply(Delay[sapply(Delay, is.character)], as.factor)

## Combination of ArrDelay and DepDelay into TotalDelay
TotalDelay <- as.matrix(rowSums(seven[,c("ArrDelay", "DepDelay")], na.rm=TRUE))
Delay <- cbind(Delay, TotalDelay)
Delay$TotalDelay <- as.integer(Delay$TotalDelay)

## Removing all the cases where variables read NA as it might interfere with the regression calculations
Delay <- Delay[complete.cases(Delay),]
DelaySet <- Delay[,c(6, 8, 13, 17, 28)]

x <- data.matrix(DelaySet[,1:4])
y <- as.matrix(DelaySet[,5])

# 70-30 Train test split
set.seed(2022)
train <- sample.split(Y = Delay$TotalDelay , SplitRatio = 0.7)
trainset1 <- subset(DelaySet, train == T)
testset1 <- subset(DelaySet, train == F)

mlr1 <- lm(TotalDelay ~ ., data=trainset1)
summary(mlr1)
rmse <- mean(residuals(mlr1)^2)
rmse <- sqrt(rmse)
rmse
## 73.56909

## The multiple linear regression model for Delay can be improved via lasso/ridge
DelaySetA <- data.matrix(DelaySet)
trainset2 <- subset(DelaySetA, train == T)
testset2 <- subset(DelaySetA, train == F)
```
## 73.56909, RMSE can be improved via lasso/ridge

## Ridge Regression

```{r}
grid <- 10^seq(10,-2,length=25)
## Since its a big dataset, i will only be using the minimum amount of folds
ridge.mod <- cv.glmnet(x = trainset2[, 1:4], y = trainset2[,5], alpha = 0, lambda = grid)
cv.out = cv.glmnet(x = trainset2[, 1:4], y = trainset2[,5], alpha = 0)
bestlambda=cv.out$lambda.min
bestlambda
## 1.07958
ridge.prediction=predict(ridge.mod, s= bestlambda, newx = testset2[, 1:4])

## Summary of ridge predictions
summary(ridge.prediction)
RMSE.ridge <- sqrt(mean((ridge.prediction-testset2[,5])^2))
RMSE.ridge
# 72.21144
outcome <- glmnet(x,y,alpha=0, lambda=grid)
ridge.coef <- predict(outcome,type="coefficients",s=bestlambda)
ridge.coef
head(ridge.prediction, 5)
```
## best lambda = 1.07958, Rmse = 72.21144

## Lasso Regression

```{r}
lasso.mod <- cv.glmnet(x = trainset2[, 1:4], y = trainset2[,5], alpha = 1, lambda = grid)
cv.out = cv.glmnet(x = trainset2[, 1:4], y = trainset2[,5], alpha = 1)
bestlambdal=cv.out$lambda.min
bestlambdal
## 0.004783209
lasso.prediction=predict(lasso.mod, s = bestlambda, newx = testset2[,1:4])

## Summary of lasso predictions
summary(lasso.prediction)
RMSE.lasso <- sqrt(mean((lasso.prediction-testset2[,5])^2))
RMSE.lasso
# 72.26279
outcome2 <- glmnet(x,y,alpha=1, lambda=grid)
lasso.coef <- predict(outcome2,type="coefficients",s=bestlambda)
lasso.coef
head(lasso.prediction)
```
## best lambda = 0.004783209, Rmse = 72.26279
## Since Ridge regression has a lower RMSE as compared to Lasso regression, the ridge regression model will be used to represent the prediction model

## Logistic Regression

```{r}
## Changing the variable TotalDelay into a binary one where if the delay is more than 10 minutes, it'll be 1 and 0 otherwise
library(caret)
library(dplyr)

DelaySetL <- DelaySet
DelaySetL <- DelaySetL %>%
  mutate(Delayed = if_else(TotalDelay > 10, 1 , 0))
DelaySetL$Delayed <- as.factor(DelaySetL$Delayed)

## TotalDelay variable is removed as it is now represented as Delayed
DelaySetL <- DelaySetL %>% select(-TotalDelay)
trainset3 <- subset(DelaySetL, train == T)
testset3 <- subset(DelaySetL, train == F)

logreg.mod <- train(Delayed ~ ., data = trainset3, method = "glm", family = "binomial",
                    trControl=trainControl(method = "cv", number = 3))

logreg.predict <- predict(logreg.mod, testset3)

confusionmatrix <- confusionMatrix(logreg.predict, testset3[,"Delayed"])
confusionmatrix
# (1391065+39251)/(2182528) = 0.655 (3s.f)
# Accuracy of 65.5%
```
# (1391065+39251)/(2182528) = 0.655 (3s.f)
# Accuracy of 65.5%